---
title: "Brussels health-precarity map"
author: Médor
---

::: {.callout-note appearance="simple"}
## Credit
This recreates the building-level map from *Bruxelles malade*, a data-journalism
investigation by the Belgian cooperative magazine **[Médor](https://bxl-malade.medor.coop/)**.
The original is a scroll-driven D3/Mapbox map; here it is rebuilt in R from the
**same underlying data** (the article's own building-footprint and per-sector
data files). All credit for the reporting and design goes to Médor.
:::

The original shades Brussels' **building footprints** by a health-precarity
index (`dim`) on a dark basemap. Since the piece is not written in R, I extracted
its merged-buildings geometry (721 dissolved footprints, one per statistical
sector, each carrying its `dim` value) and rebuilt the map with `sf` and
`ggplot2`, reusing Médor's exact color ramp. Higher values (red) mean more
precarious; lower values (teal) mean better off. **Hover over a district** to
highlight it and see its area-weighted average index (rendered interactively
with `ggiraph`).

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(sf)
library(dplyr)
library(ggplot2)
library(scales)
library(ggiraph)
```

```{r}
#| label: choropleth
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8.5
#| fig-height: 9
crs_local <- 3812 # Belgian Lambert 2008

# Building footprints, each carrying the precarity index `dim`
buildings <- read_sf(file.path("choropleth-data", "bxlBuildings.topo.json")) |>
  st_set_crs(4326) |>
  st_transform(crs_local) |>
  # the sector code's first 5 digits are the commune (NIS) code
  mutate(commune_code = substr(sector, 1, 5),
         area = as.numeric(st_area(geometry)))

# Commune outlines + labels, dissolved from the statistical sectors
sectors <- read_sf(file.path("choropleth-data", "brusselsSectors.json")) |>
  st_set_crs(4326) |>
  st_transform(crs_local) |>
  st_make_valid() |>
  mutate(commune_code = substr(cd_sector, 1, 5))

communes <- sectors |>
  group_by(commune_code, commune = tx_munty_descr_fr) |>
  summarise(.groups = "drop")

# Area-weighted average index per district (commune)
district_dim <- buildings |>
  st_drop_geometry() |>
  filter(!is.na(dim)) |>
  group_by(commune_code) |>
  summarise(dim_district = weighted.mean(dim, area), .groups = "drop")

communes <- communes |>
  left_join(district_dim, by = "commune_code") |>
  mutate(tooltip = sprintf("%s\nPrecarity index: %.1f", commune, dim_district),
         label_pt = st_point_on_surface(geometry))

# Médor's diverging ramp with the article's exact break points (dim 0 -> 50)
stops <- c(0, 11, 14, 18, 24, 29, 34, 41, 50)
cols  <- c("#00939C", "#49A6AE", "#65B3BA", "#B2DCDF", "#FEEEE8",
           "#F8C0AA", "#EC9370", "#DA6436", "#C22E00")

bg <- "#1b1b1b"

p <- ggplot() +
  geom_sf(data = buildings, aes(fill = dim), color = NA) +
  # interactive district layer: transparent until hovered
  geom_sf_interactive(
    data = communes,
    aes(tooltip = tooltip, data_id = commune_code),
    fill = "transparent", color = "white", linewidth = 0.3
  ) +
  geom_sf_text(data = communes, aes(label = commune, geometry = label_pt),
               color = "grey85", size = 2.6, fontface = "bold", check_overlap = TRUE) +
  scale_fill_gradientn(
    colours = cols,
    values = rescale(stops, from = c(0, 50)),
    limits = c(0, 50), oob = squish,
    na.value = "#3a3a3a",
    name = "Health-precarity index",
    guide = guide_colourbar(barwidth = unit(45, "mm"), barheight = unit(4, "mm"),
                            title.position = "top", ticks = FALSE)
  ) +
  labs(
    title = "A divided city",
    subtitle = "Building footprints shaded by Médor's index — hover a district for its average",
    caption = "Data extracted from Médor, 'Bruxelles malade'."
  ) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = bg, color = NA),
    panel.background = element_rect(fill = bg, color = NA),
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_text(color = "grey85", size = 9),
    legend.text = element_text(color = "grey85", size = 8),
    plot.title = element_text(face = "bold", color = "white", margin = margin(b = 2)),
    plot.subtitle = element_text(color = "grey70", size = 11, margin = margin(b = 8)),
    plot.caption = element_text(color = "grey55", size = 8, hjust = 0, margin = margin(t = 8)),
    plot.margin = margin(12, 12, 12, 12)
  )

girafe(
  ggobj = p,
  width_svg = 8.5, height_svg = 9,
  options = list(
    opts_hover(css = "fill:#ffffff55;stroke:white;stroke-width:1.5px;"),
    opts_tooltip(css = "background:#1b1b1b;color:white;padding:6px 10px;border-radius:4px;font-family:sans-serif;font-size:12px;white-space:pre;"),
    opts_toolbar(saveaspng = FALSE),
    opts_sizing(rescale = TRUE)
  )
)
```

The pattern is the article's central finding: a "poor croissant" of high-index
(red) neighbourhoods curls through the inner-city and industrial west
(Molenbeek, Anderlecht, Saint-Josse), while the wealthier south-east
(Uccle, Woluwe) sits at the teal end. The extracted per-sector data also carries
the other indicators used in the piece (air pollution `no2`/`pm25`, green
coverage, noise, diabetes rates, cancer-screening uptake), so the same footprints
can be recolored by any of them.
```{r}
#| ref.label: ["setup", "choropleth"]
#| eval: false
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
```

## How it is built, step by step

Each tab adds one layer of the recipe on top of the previous one, so you can see
what every piece of the final map contributes. The code for each step is shown
below its plot.

::: {.panel-tabset}

### 1. The geometry

The extracted footprints are one dissolved polygon per statistical sector, each
carrying its precarity index. Reprojecting to Belgian Lambert 2008 (EPSG:3812)
is what lets areas be measured in metres for the weighted averages later.

```{r}
#| label: cstep-1
#| echo: false
#| message: false
#| warning: false
buildings |>
  select(sector, commune_code, dim, area) |>
  head(5)
```

```{r}
#| ref.label: cstep-1
#| eval: false
#| echo: true
```

### 2. Footprints

A single `geom_sf()` with no outline colour. At this zoom the footprints read as
texture rather than individual buildings, which is the effect the original is
after.

```{r}
#| label: cstep-2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 7
m2 <- ggplot() +
  geom_sf(data = buildings, aes(fill = dim), color = NA)
m2
```

```{r}
#| ref.label: cstep-2
#| eval: false
#| echo: true
```

### 3. Médor's colour ramp

The default blue gradient hides the story. `scale_fill_gradientn()` with the
article's nine stops and colours puts the neutral cream tone at the middle of
the index, so teal and red read as "better off" and "more precarious".

```{r}
#| label: cstep-3
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 7
m3 <- m2 +
  scale_fill_gradientn(
    colours = cols,
    values = rescale(stops, from = c(0, 50)),
    limits = c(0, 50), oob = squish,
    na.value = "#3a3a3a",
    name = "Health-precarity index",
    guide = guide_colourbar(barwidth = unit(45, "mm"), barheight = unit(4, "mm"),
                            title.position = "top", ticks = FALSE)
  )
m3
```

```{r}
#| ref.label: cstep-3
#| eval: false
#| echo: true
```

### 4. District outlines

The commune boundaries and their names give the reader somewhere to stand. The
outline layer is `geom_sf_interactive()` with a transparent fill: it is invisible
until hovered, when it becomes the highlight and tooltip target.

```{r}
#| label: cstep-4
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 7
m4 <- m3 +
  geom_sf_interactive(
    data = communes,
    aes(tooltip = tooltip, data_id = commune_code),
    fill = "transparent", color = "white", linewidth = 0.3
  ) +
  geom_sf_text(data = communes, aes(label = commune, geometry = label_pt),
               color = "grey85", size = 2.6, fontface = "bold", check_overlap = TRUE)
m4
```

```{r}
#| ref.label: cstep-4
#| eval: false
#| echo: true
```

### 5. Dark theme

`theme_void()` strips the axes and graticule, and the near-black background is
what makes the ramp glow the way the original does. The legend sits top-left,
above the map.

```{r}
#| label: cstep-5
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 7.5
m5 <- m4 +
  labs(
    title = "A divided city",
    subtitle = "Building footprints shaded by Médor's index",
    caption = "Data extracted from Médor, 'Bruxelles malade'."
  ) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 13) +
  theme(
    plot.background = element_rect(fill = bg, color = NA),
    panel.background = element_rect(fill = bg, color = NA),
    legend.position = "top",
    legend.justification = "left",
    legend.title = element_text(color = "grey85", size = 9),
    legend.text = element_text(color = "grey85", size = 8),
    plot.title = element_text(face = "bold", color = "white", margin = margin(b = 2)),
    plot.subtitle = element_text(color = "grey70", size = 11, margin = margin(b = 8)),
    plot.caption = element_text(color = "grey55", size = 8, hjust = 0, margin = margin(t = 8)),
    plot.margin = margin(12, 12, 12, 12)
  )
m5
```

```{r}
#| ref.label: cstep-5
#| eval: false
#| echo: true
```

### 6. Interactivity

Nothing above is interactive yet: it is a static ggplot with interactive
*aesthetics* recorded on the district layer. `ggiraph::girafe()` renders it to
SVG and wires those up, with the hover CSS supplying the white wash.

```{r}
#| label: cstep-6
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8.5
#| fig-height: 9
girafe(
  ggobj = m5,
  width_svg = 8.5, height_svg = 9,
  options = list(
    opts_hover(css = "fill:#ffffff55;stroke:white;stroke-width:1.5px;"),
    opts_tooltip(css = "background:#1b1b1b;color:white;padding:6px 10px;border-radius:4px;font-family:sans-serif;font-size:12px;white-space:pre;"),
    opts_toolbar(saveaspng = FALSE),
    opts_sizing(rescale = TRUE)
  )
)
```

```{r}
#| ref.label: cstep-6
#| eval: false
#| echo: true
```

:::
