---
title: "Ridgeline plot with inside plot and annotations"
author: Ansgar Wolsing
---

::: {.callout-note appearance="simple"}
## Credit
This reproduces a visualization created by **Ansgar Wolsing** for TidyTuesday,
as featured on the [R Graph Gallery](https://r-graph-gallery.com/web-ridgeline-plot-with-inside-plot-and-annotations.html).
Data: Kate Pennington (2018), *Bay Area Craigslist Rental Housing Posts,
2000-2018*. All credit for the original design goes to the author.
:::

A distribution-per-adjective ridgeline of San Francisco Bay Area Craigslist
rental prices, with an annotated "legend" plot embedded inside the panel.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggtext)
library(ggdist)
library(glue)
library(patchwork)
library(MetBrewer)
library(showtext)

# Fira Sans is not a system font here; pull it from Google via showtext.
font_add_google("Fira Sans", "Fira Sans")
font_add_google("Fira Sans", "Fira Sans SemiBold")
showtext_auto()
```

```{r}
#| label: ridgeline
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 11
#| fig-dpi: 150
# Load datasets (slimmed copies of the originals)
df_plot <- read_csv(file.path("ridgeline-data", "df_plot.csv"))
rent_title_words <- read_csv(file.path("ridgeline-data", "rent_legend.csv"))
rent_stats <- read_csv(file.path("ridgeline-data", "rent_stats.csv"))

# Prepare data
df_plot <- df_plot %>% arrange(desc(mean_price))
df_plot$word <- factor(df_plot$word, levels = unique(df_plot$word))

# Statistics (precomputed from the full rent dataset)
mean_price <- rent_stats$mean_price
median_price <- rent_stats$median_price
n_rental_posts <- rent_stats$n_rental_posts

bg_color <- "grey97"
font_family <- "Fira Sans"

plot_subtitle = glue("Adjectives used to describe houses and apartments
in the San Francisco Bay Area in the titles of rental posts on Craigslist and
how they are related to rental prices. Titles from
{scales::number(n_rental_posts, big.mark = ',')} rental posts on Craigslist
between 2000 and 2018.
The 15 most frequent adjectives are shown.
")

# Main plot
p <- df_plot %>%
  ggplot(aes(word, price)) +
  stat_halfeye(fill_type = "segments", alpha = 0.3) +
  stat_interval() +
  stat_summary(geom = "point", fun = median) +
  annotate("text", x = 16, y = 0, label = "(\U00F8 bedrooms)",
           family = "Fira Sans", size = 3, hjust = 0.5) +
  stat_summary(
    aes(y = beds),
    geom = "text",
    fun.data = function(x) {
      data.frame(
        y = 0,
        label = sprintf("(%s)", scales::number(mean(ifelse(x > 0, x, NA), na.rm = TRUE), accuracy = 0.1)))},
    family = font_family, size = 2.5
  ) +
  geom_hline(yintercept = median_price, col = "grey30", lty = "dashed") +
  annotate("text", x = 16, y = median_price + 50, label = "Median rent",
           family = "Fira Sans", size = 3, hjust = 0) +
  scale_x_discrete(labels = toupper) +
  scale_y_continuous(breaks = seq(2500, 20000, 2500)) +
  scale_color_manual(values = MetBrewer::met.brewer("VanGogh3")) +
  coord_flip(ylim = c(0, 10000), clip = "off") +
  guides(col = "none") +
  labs(
    title = toupper("Nice and Clean - Relatively Low Rent?"),
    subtitle = plot_subtitle,
    caption = "Axis capped at 10,000 USD.<br>
    Data: Pennington, Kate (2018).
    Bay Area Craigslist Rental Housing Posts, 2000-2018.<br>
    Retrieved from github.com/katepennington/historic_bay_area_craigslist_housing_posts/blob/master/clean_2000_2018.csv.zip.
    <br>
    Visualization: Ansgar Wolsing",
    x = NULL,
    y = "Rent in USD"
  ) +
  theme_minimal(base_family = font_family) +
  theme(
    plot.background = element_rect(color = NA, fill = bg_color),
    panel.grid = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.1, color = "grey75"),
    plot.title = element_text(family = "Fira Sans SemiBold"),
    plot.title.position = "plot",
    plot.subtitle = element_textbox_simple(
      margin = margin(t = 4, b = 16), size = 10),
    plot.caption = element_textbox_simple(
      margin = margin(t = 12), size = 7
    ),
    plot.caption.position = "plot",
    axis.text.y = element_text(hjust = 0, margin = margin(r = -10), family = "Fira Sans SemiBold"),
    plot.margin = margin(4, 4, 4, 4)
  )

# Create legend (inside plot)
df_for_legend <- rent_title_words %>%
  filter(word == "beautiful")

p_legend <- df_for_legend %>%
  ggplot(aes(word, price)) +
  stat_halfeye(fill_type = "segments", alpha = 0.3) +
  stat_interval() +
  stat_summary(geom = "point", fun = median) +
  annotate(
    "richtext",
    x = c(0.8, 0.8, 0.8, 1.4, 1.8),
    y = c(1000, 5000, 3000, 2400, 4000),
    label = c("50 % of prices<br>fall within this range", "95 % of prices",
              "80 % of prices", "Median", "Distribution<br>of prices"),
    fill = NA, label.size = 0, family = font_family, size = 2, vjust = 1,
  ) +
  geom_curve(
    data = data.frame(
      x = c(0.7, 0.80, 0.80, 1.225, 1.8),
      xend = c(0.95, 0.95, 0.95, 1.075, 1.8),
      y = c(1800, 5000, 3000, 2300, 3800),
      yend = c(1800, 5000, 3000, 2100, 2500)),
    aes(x = x, xend = xend, y = y, yend = yend),
    stat = "unique", curvature = 0.2, size = 0.2, color = "grey12",
    arrow = arrow(angle = 20, length = unit(1, "mm"))
  ) +
  scale_color_manual(values = MetBrewer::met.brewer("VanGogh3")) +
  coord_flip(xlim = c(0.75, 1.3), ylim = c(0, 6000), expand = TRUE) +
  guides(color = "none") +
  labs(title = "Legend") +
  theme_void(base_family = font_family) +
  theme(plot.title = element_text(family = "Fira Sans SemiBold", size = 9,
                                  hjust = 0.075),
        plot.background = element_rect(color = "grey30", size = 0.2, fill = bg_color))

# Combine main plot with embedded legend
p + inset_element(p_legend, l = 0.6, r = 1.0, t = 0.99, b = 0.7, clip = FALSE)
```

```{r}
#| ref.label: ["setup", "ridgeline"]
#| eval: false
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
```

## How it is built, step by step

Each tab adds one layer of the recipe on top of the previous one, so you can see
what every piece of the final chart contributes. The code for each step is shown
below its plot.

::: {.panel-tabset}

### 1. Distributions

`ggdist::stat_halfeye()` draws a density curve per adjective. `coord_flip()` is
applied from the start so the orientation stays the same across all steps.

```{r}
#| label: step-1
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step1 <- ggplot(df_plot, aes(word, price)) +
  stat_halfeye(fill_type = "segments", alpha = 0.3) +
  coord_flip(ylim = c(0, 10000), clip = "off")
step1
```

```{r}
#| ref.label: step-1
#| eval: false
#| echo: true
```

### 2. Intervals

`stat_interval()` adds the nested 50 / 80 / 95 % interval bars, and
`stat_summary()` marks the median of each group with a point.

```{r}
#| label: step-2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step2 <- ggplot(df_plot, aes(word, price)) +
  stat_halfeye(fill_type = "segments", alpha = 0.3) +
  stat_interval() +
  stat_summary(geom = "point", fun = median) +
  coord_flip(ylim = c(0, 10000), clip = "off")
step2
```

```{r}
#| ref.label: step-2
#| eval: false
#| echo: true
```

### 3. Colour

The three interval widths get their colours from
`MetBrewer::met.brewer("VanGogh3")`, and the default colour legend is dropped
because the inset legend explains it instead.

```{r}
#| label: step-3
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step3 <- step2 +
  scale_color_manual(values = MetBrewer::met.brewer("VanGogh3")) +
  guides(col = "none")
step3
```

```{r}
#| ref.label: step-3
#| eval: false
#| echo: true
```

### 4. Reference line

A dashed `geom_hline()` at the overall median rent gives every distribution a
common anchor to be read against.

```{r}
#| label: step-4
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step4 <- step3 +
  geom_hline(yintercept = median_price, col = "grey30", lty = "dashed") +
  annotate("text", x = 16, y = median_price + 50, label = "Median rent",
           family = font_family, size = 3, hjust = 0)
step4
```

```{r}
#| ref.label: step-4
#| eval: false
#| echo: true
```

### 5. Bedroom counts

A text-producing `stat_summary()` writes the mean number of bedrooms for each
adjective at the left edge, which is what keeps the price differences
interpretable.

```{r}
#| label: step-5
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step5 <- step4 +
  annotate("text", x = 16, y = 0, label = "(\U00F8 bedrooms)",
           family = font_family, size = 3, hjust = 0.5) +
  stat_summary(
    aes(y = beds),
    geom = "text",
    fun.data = function(x) {
      data.frame(
        y = 0,
        label = sprintf("(%s)", scales::number(mean(ifelse(x > 0, x, NA), na.rm = TRUE), accuracy = 0.1)))},
    family = font_family, size = 2.5
  )
step5
```

```{r}
#| ref.label: step-5
#| eval: false
#| echo: true
```

### 6. Scales and labels

Axis breaks every 2,500 USD, upper-case category labels, and the title,
subtitle and caption text.

```{r}
#| label: step-6
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 7
#| fig-dpi: 150
step6 <- step5 +
  scale_x_discrete(labels = toupper) +
  scale_y_continuous(breaks = seq(2500, 20000, 2500)) +
  labs(
    title = toupper("Nice and Clean - Relatively Low Rent?"),
    subtitle = plot_subtitle,
    caption = "Axis capped at 10,000 USD.<br>Visualization: Ansgar Wolsing",
    x = NULL,
    y = "Rent in USD"
  )
step6
```

```{r}
#| ref.label: step-6
#| eval: false
#| echo: true
```

### 7. Theme

`theme_minimal()` plus a light background, vertical-only grid lines, left-aligned
category labels and `ggtext` text boxes that wrap the subtitle and caption.

```{r}
#| label: step-7
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 8
#| fig-dpi: 150
step7 <- step6 +
  theme_minimal(base_family = font_family) +
  theme(
    plot.background = element_rect(color = NA, fill = bg_color),
    panel.grid = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.1, color = "grey75"),
    plot.title = element_text(family = "Fira Sans SemiBold"),
    plot.title.position = "plot",
    plot.subtitle = element_textbox_simple(margin = margin(t = 4, b = 16), size = 10),
    plot.caption = element_textbox_simple(margin = margin(t = 12), size = 7),
    plot.caption.position = "plot",
    axis.text.y = element_text(hjust = 0, margin = margin(r = -10),
                              family = "Fira Sans SemiBold"),
    plot.margin = margin(4, 4, 4, 4)
  )
step7
```

```{r}
#| ref.label: step-7
#| eval: false
#| echo: true
```

### 8. Inset legend

The annotated legend is a second, standalone plot built from a single adjective
(`beautiful`) with `annotate("richtext", ...)` labels and `geom_curve()` arrows.

```{r}
#| label: step-8
#| echo: false
#| message: false
#| warning: false
#| fig-width: 5
#| fig-height: 3
#| fig-dpi: 150
p_legend
```

```{r}
#| ref.label: step-8
#| eval: false
#| echo: true
```

### 9. Combined

`patchwork::inset_element()` places the legend in the empty space at the top
right of the main panel.

```{r}
#| label: step-9
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 11
#| fig-dpi: 150
step7 + inset_element(p_legend, l = 0.6, r = 1.0, t = 0.99, b = 0.7, clip = FALSE)
```

```{r}
#| ref.label: step-9
#| eval: false
#| echo: true
```

:::
