---
title: "Subway table (gt)"
---

::: {.callout-note appearance="simple"}
## Credit
This is a rebuild of a table originally created by **Karina Bartolomé**:
[*The grammar of tables*](https://karbartolome.quarto.pub/the-grammar-of-tables/)
· [source code](https://github.com/karbartolome/quartopub/tree/main/01_tables_r_python).
All credit for the original design goes to the author.
:::

It uses the **real Buenos Aires subway geography** (station coordinates and the
city's *barrios* polygons) with **placeholder passenger numbers**, and brings
back the per-line inline graphics from the original: a polar **clock plot** of
hourly use, a **map** with station points sized by passenger share, a stacked
time-of-day bar, and a daily **evolution** line.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(gt)
library(gtExtras)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(sf)

here <- function(...) file.path("data", ...)

# Buenos Aires subway lines and their official colors
line_colors <- c(
  A = "#18cccc", B = "#eb0909", C = "#233aa8",
  D = "#02db2e", E = "#c618cc", H = "#ffdd00"
)

# --- Real geography -------------------------------------------------------
map_caba <- st_read(here("barrios.geojson"), quiet = TRUE)

stations <- readr::read_csv(here("stations.csv"), show_col_types = FALSE) |>
  rename(line = linea, station = estacion) |>
  filter(line %in% names(line_colors))

set.seed(42)

# --- Placeholder passenger data -------------------------------------------
# Per-station total passengers for the month
stations <- stations |>
  mutate(pax_station = round(runif(n(), 5e4, 6e5)))

# Hourly distribution per line (a plausible double-peaked commuter shape)
base_hour <- dnorm(0:23, 8, 2) + 0.8 * dnorm(0:23, 18, 2.5) + 0.05
hourly <- map_dfr(names(line_colors), function(ln) {
  w <- base_hour * runif(24, 0.7, 1.3)
  tibble(line = ln, hour = 0:23, pax_total = round(w / sum(w) * 4e6))
})

# Daily evolution per line across the month
daily <- map_dfr(names(line_colors), function(ln) {
  d <- seq(as.Date("2021-11-01"), as.Date("2021-11-30"), by = "day")
  wk <- ifelse(format(d, "%u") %in% c("6", "7"), 0.45, 1)
  tibble(line = ln, date = d,
         pax_total = round(wk * runif(length(d), 8e4, 1.3e5)))
})

# Circular mean hour, weighted by passengers (replaces the {circular} pkg)
mean_hours <- hourly |>
  group_by(line) |>
  summarise(
    mean_hour = {
      ang <- hour / 24 * 2 * pi
      m <- atan2(sum(pax_total * sin(ang)), sum(pax_total * cos(ang)))
      (m / (2 * pi) * 24) %% 24
    },
    .groups = "drop"
  )

# --- Plot builders (ported from the original R code) ----------------------
fig_clock_plot <- function(.line) {
  mean_hour <- mean_hours$mean_hour[mean_hours$line == .line]
  temp <- hourly |>
    filter(line == .line) |>
    mutate(color_hour = hour == round(mean_hour))
  ggplot(temp, aes(hour, pax_total)) +
    geom_col(color = "white", fill = "lightgrey") +
    coord_polar(start = 0) +
    geom_vline(xintercept = mean_hour, color = line_colors[[.line]], linewidth = 2) +
    geom_label(
      aes(y = max(pax_total) + quantile(pax_total, 0.3),
          color = color_hour, label = hour),
      size = 6, label.size = NA, show.legend = FALSE
    ) +
    scale_color_manual(values = c("black", line_colors[[.line]])) +
    scale_x_continuous("", limits = c(0, 24), breaks = 0:24) +
    scale_y_continuous(labels = scales::unit_format(unit = "K", scale = 1e-3)) +
    labs(y = NULL) +
    theme_minimal() +
    theme(text = element_text(size = 22, color = "grey"),
          axis.text.x = element_blank(), panel.grid = element_blank())
}

fig_map <- function(.line) {
  temp <- stations |>
    filter(line == .line) |>
    mutate(pax_percent = pax_station / sum(pax_station))
  ggplot() +
    geom_sf(data = map_caba, color = "black", fill = "white", linewidth = 0.1) +
    geom_point(
      data = temp, aes(long, lat, size = pax_percent),
      alpha = 0.7, fill = line_colors[[.line]], color = "black", shape = 21
    ) +
    scale_size_continuous(range = c(1, 9), guide = "none") +
    theme_void()
}

fig_evol <- function(.line) {
  ggplot(filter(daily, line == .line), aes(date, pax_total)) +
    geom_line(color = "grey", linewidth = 2.5) +
    geom_line(color = line_colors[[.line]], linewidth = 1.5) +
    scale_y_continuous(labels = scales::unit_format(unit = "K", scale = 1e-3)) +
    labs(x = NULL, y = NULL) +
    theme_minimal() +
    theme(text = element_text(size = 22),
          panel.grid = element_blank(), axis.text.x = element_blank())
}

# --- One row per line -----------------------------------------------------
tbl_data <- stations |>
  group_by(line) |>
  summarise(
    most_used_station = station[which.max(pax_station)],
    pax_total = sum(pax_station),
    .groups = "drop"
  ) |>
  mutate(
    variation = runif(n(), -0.12, 0.18),
    route = c(
      "Plaza de Mayo - San Pedrito", "Alem - J. M. de Rosas",
      "Retiro - Constitución", "Catedral - Congreso de Tucumán",
      "Bolívar - Plaza de los Virreyes", "Facultad de Derecho - Hospitales"
    ),
    # placeholder text columns; ggplots are injected via text_transform below
    clock_plot = "", passengers_per_day = "", map = ""
  ) |>
  rowwise() |>
  mutate(passengers_type = list(as.numeric(round(prop.table(runif(3)) * 100)))) |>
  ungroup() |>
  select(line, route, most_used_station, clock_plot, pax_total, variation,
         passengers_type, passengers_per_day, map)

# ggplot objects held separately, in table row order
clock_plots <- map(tbl_data$line, fig_clock_plot)
evol_plots   <- map(tbl_data$line, fig_evol)
map_plots    <- map(tbl_data$line, fig_map)
```

```{r}
#| label: table
#| echo: false
#| message: false
#| warning: false
#| column: page
tbl_data |>
  gt() |>
  tab_header(
    title = md("**A summary of subway use in Buenos Aires**"),
    subtitle = "Real geography, placeholder passenger data (styled after 'The grammar of tables')"
  ) |>
  opt_align_table_header("left") |>
  tab_spanner(label = "Use per hour", columns = clock_plot) |>
  tab_spanner(label = "Number of passengers", columns = c(pax_total, variation)) |>
  cols_align("center", columns = c(pax_total, variation, most_used_station)) |>
  cols_width(
    line ~ px(50), route ~ px(150), most_used_station ~ px(110),
    clock_plot ~ px(150), pax_total ~ px(80), variation ~ px(80),
    passengers_type ~ px(120), passengers_per_day ~ px(150), map ~ px(120)
  ) |>
  # color each line label
  {\(x) Reduce(function(tbl, ln) {
    tab_style(tbl, cell_text(color = line_colors[[ln]], weight = "bold"),
              cells_body(columns = line, rows = line == ln))
  }, names(line_colors), x)}() |>
  tab_style(cell_fill(color = "#f0f0f0"), cells_body(columns = route)) |>
  fmt_number(pax_total, suffixing = TRUE) |>
  fmt_percent(variation, decimals = 1) |>
  text_transform(
    locations = cells_body(columns = clock_plot),
    fn = function(x) map(clock_plots, ggplot_image, height = px(120), aspect_ratio = 1)
  ) |>
  text_transform(
    locations = cells_body(columns = passengers_per_day),
    fn = function(x) map(evol_plots, ggplot_image, height = px(90), aspect_ratio = 2)
  ) |>
  text_transform(
    locations = cells_body(columns = map),
    fn = function(x) map(map_plots, ggplot_image, height = px(120), aspect_ratio = 0.9)
  ) |>
  gt_plt_bar_stack(
    column = passengers_type, position = "fill",
    labels = c("Morning", "Afternoon", "Night"),
    palette = c("grey", "#A3B1C9", "#4C699E"),
    fmt_fn = scales::label_percent(scale = 1), width = 60
  ) |>
  cols_label(
    line = "Line", route = "Route", most_used_station = "Most used station",
    clock_plot = md("Passengers per hour"), pax_total = md("Total"),
    variation = md("% Variation"), passengers_per_day = md("Passengers per day"),
    map = md("% Passengers per station")
  ) |>
  tab_source_note(md(paste0(
    "Placeholder rebuild of a table by Karina Bartolomé. Geography: Open Data ",
    "Portal, Autonomous City of Buenos Aires; passenger figures are simulated."
  ))) |>
  tab_options(
    data_row.padding = px(2),
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden"
  )
```

```{r}
#| ref.label: ["setup", "table"]
#| eval: false
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
```

## How it is built, step by step

Each tab adds one piece of the recipe on top of the previous one, so you can see
what every part of the finished table contributes. The code for each step is
shown below its output.

::: {.panel-tabset}

### 1. One row per line

Everything starts from a six-row summary, one row per subway line. The
`clock_plot`, `passengers_per_day` and `map` columns are deliberately empty
strings: they are placeholders whose contents get swapped for images later.

```{r}
#| label: tstep-1
#| echo: false
#| message: false
#| warning: false
tbl_data |>
  select(line, route, most_used_station, pax_total, variation)
```

```{r}
#| ref.label: tstep-1
#| eval: false
#| echo: true
```

### 2. The inline plots

Each graphic is an ordinary standalone ggplot, built by a function that takes a
line name. They are stripped down hard (no axis text, no legend) because they
will be rendered a few dozen pixels tall. Here are all three for line A.

```{r}
#| label: tstep-2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 3.5
library(patchwork)
fig_clock_plot("A") + fig_evol("A") + fig_map("A")
```

```{r}
#| ref.label: tstep-2
#| eval: false
#| echo: true
```

### 3. A bare gt table

`gt()` on its own already gives a competent table. Everything from here is
narrowing what the reader has to work to see.

```{r}
#| label: tstep-3
#| echo: false
#| message: false
#| warning: false
#| column: page
t3 <- tbl_data |>
  select(-passengers_type) |>
  gt()
t3
```

```{r}
#| ref.label: tstep-3
#| eval: false
#| echo: true
```

### 4. Structure

A left-aligned title, spanners that group related columns under one heading, and
explicit column widths. Fixing the widths matters here because the image columns
have no natural size of their own.

```{r}
#| label: tstep-4
#| echo: false
#| message: false
#| warning: false
#| column: page
t4 <- tbl_data |>
  gt() |>
  tab_header(
    title = md("**A summary of subway use in Buenos Aires**"),
    subtitle = "Real geography, placeholder passenger data (styled after 'The grammar of tables')"
  ) |>
  opt_align_table_header("left") |>
  tab_spanner(label = "Use per hour", columns = clock_plot) |>
  tab_spanner(label = "Number of passengers", columns = c(pax_total, variation)) |>
  cols_align("center", columns = c(pax_total, variation, most_used_station)) |>
  cols_width(
    line ~ px(50), route ~ px(150), most_used_station ~ px(110),
    clock_plot ~ px(150), pax_total ~ px(80), variation ~ px(80),
    passengers_type ~ px(120), passengers_per_day ~ px(150), map ~ px(120)
  )
t4
```

```{r}
#| ref.label: tstep-4
#| eval: false
#| echo: true
```

### 5. Colour and number formatting

Each line label is painted in that line's official colour by folding
`tab_style()` over the six lines, the route column gets a light fill to set it
apart from the numbers, and the totals and variations are formatted as
suffixed numbers and percentages.

```{r}
#| label: tstep-5
#| echo: false
#| message: false
#| warning: false
#| column: page
t5 <- t4 |>
  {\(x) Reduce(function(tbl, ln) {
    tab_style(tbl, cell_text(color = line_colors[[ln]], weight = "bold"),
              cells_body(columns = line, rows = line == ln))
  }, names(line_colors), x)}() |>
  tab_style(cell_fill(color = "#f0f0f0"), cells_body(columns = route)) |>
  fmt_number(pax_total, suffixing = TRUE) |>
  fmt_percent(variation, decimals = 1)
t5
```

```{r}
#| ref.label: tstep-5
#| eval: false
#| echo: true
```

### 6. Injecting the plots

`text_transform()` replaces the contents of each placeholder column with images
rendered from the ggplot lists via `gt::ggplot_image()`. The lists are in table
row order, which is the only thing keeping plots matched to their line.

```{r}
#| label: tstep-6
#| echo: false
#| message: false
#| warning: false
#| column: page
t6 <- t5 |>
  text_transform(
    locations = cells_body(columns = clock_plot),
    fn = function(x) map(clock_plots, ggplot_image, height = px(120), aspect_ratio = 1)
  ) |>
  text_transform(
    locations = cells_body(columns = passengers_per_day),
    fn = function(x) map(evol_plots, ggplot_image, height = px(90), aspect_ratio = 2)
  ) |>
  text_transform(
    locations = cells_body(columns = map),
    fn = function(x) map(map_plots, ggplot_image, height = px(120), aspect_ratio = 0.9)
  )
t6
```

```{r}
#| ref.label: tstep-6
#| eval: false
#| echo: true
```

### 7. Stacked bars

`gtExtras::gt_plt_bar_stack()` turns the list column of three numbers per row
into a labelled, filled bar, so the morning/afternoon/night split needs no
separate legend.

```{r}
#| label: tstep-7
#| echo: false
#| message: false
#| warning: false
#| column: page
t7 <- t6 |>
  gt_plt_bar_stack(
    column = passengers_type, position = "fill",
    labels = c("Morning", "Afternoon", "Night"),
    palette = c("grey", "#A3B1C9", "#4C699E"),
    fmt_fn = scales::label_percent(scale = 1), width = 60
  )
t7
```

```{r}
#| ref.label: tstep-7
#| eval: false
#| echo: true
```

### 8. Labels and finishing

Readable column labels, the source note that flags the passenger figures as
simulated, tight row padding, and hidden outer borders.

```{r}
#| label: tstep-8
#| echo: false
#| message: false
#| warning: false
#| column: page
t7 |>
  cols_label(
    line = "Line", route = "Route", most_used_station = "Most used station",
    clock_plot = md("Passengers per hour"), pax_total = md("Total"),
    variation = md("% Variation"), passengers_per_day = md("Passengers per day"),
    map = md("% Passengers per station")
  ) |>
  tab_source_note(md(paste0(
    "Placeholder rebuild of a table by Karina Bartolomé. Geography: Open Data ",
    "Portal, Autonomous City of Buenos Aires; passenger figures are simulated."
  ))) |>
  tab_options(
    data_row.padding = px(2),
    table.border.top.style = "hidden",
    table.border.bottom.style = "hidden"
  )
```

```{r}
#| ref.label: tstep-8
#| eval: false
#| echo: true
```

:::
