---
title: "Waffle chart for time evolution"
author: Muhammad Azhar
---

::: {.callout-note appearance="simple"}
## Credit
This reproduces a visualization created by **Muhammad Azhar** for the
#30DayChartChallenge, as featured on the
[R Graph Gallery](https://r-graph-gallery.com/web-waffle-for-time-evolution.html).
Data: NOAA (the `storms` dataset bundled with the tidyverse). All credit for the
original design goes to the author.
:::

A waffle chart showing how the yearly count of Atlantic tropical cyclones, split
by storm category, has evolved since 2010.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(showtext)
library(ggtext)
library(waffle)
library(MetBrewer)

# Load fonts
font_add_google("Outfit", "title_font")
font_add_google("Cabin", "body_font")
showtext_auto()
```

```{r}
#| label: waffle
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 9
#| fig-dpi: 320
title_font <- "title_font"
body_font <- "body_font"

# Prepare data
storms_df <- storms |>
  filter(year >= 2010,
         status %in% c("hurricane", "tropical depression",
                       "tropical storm","tropical wave")) |>
  count(year, status)

# Define text labels
title_text <- "In the Atlantic Ocean, tropical storms are more common than other cyclones."
subtitle_text <- "Tropical Storm is a cyclone with maximum sustained winds of 39 to 73 mph (34 to 63 knots).\nshowing number of cyclones for each year since 2010"
caption_text <- "Graphic: Muhammad Azhar | #30DayChartChallenge | Data: NOAA"

# Create waffle chart
ggplot(storms_df, aes(fill = status, values = n)) +
  geom_waffle(color = "white", size = .25, n_rows = 10, flip = TRUE) +
  facet_wrap(~year, nrow = 1, strip.position = "bottom") +
  scale_x_discrete() +
  scale_y_continuous(labels = function(x) x * 10,
                     expand = c(0,0)) +
  MetBrewer::scale_fill_met_d("Hiroshige", direction=1) +
  coord_equal() +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = caption_text) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(family = body_font, size=12),
    axis.text.y = element_text(family = body_font, size=12),
    legend.position = "top",
    legend.title = element_blank(),
    legend.spacing = unit(0.5, 'cm'),
    legend.key.height = unit(0.5, 'cm'),
    legend.key.width = unit(0.7, 'cm'),
    legend.text = element_text(family = body_font, size=13, face = 'plain', color = "grey10"),
    plot.title.position = "plot",
    plot.title = element_textbox(margin = margin(20, 0, 10, 0), size = 30,
                                 family = title_font, face = "bold", width = unit(55, "lines")),
    plot.subtitle = element_text(margin = margin(10, 0, 20, 0), size = 16,
                                 family = body_font, color = "grey15"),
    plot.caption = element_text(family=body_font, face="plain", size=14,
                                color="grey40", hjust=.5, margin=margin(20,0,0,0)),
    plot.background = element_rect(color="white", fill="white"),
    plot.margin = margin(20, 40, 20, 40)
  )
```

```{r}
#| ref.label: ["setup", "waffle"]
#| eval: false
#| echo: true
#| code-fold: true
#| code-summary: "Show code"
```

## How it is built, step by step

Each tab adds one layer of the recipe on top of the previous one, so you can see
what every piece of the final chart contributes. The code for each step is shown
below its plot.

::: {.panel-tabset}

### 1. Counting storms

The data work is a single `count()`: one row per year and storm status, which is
all `geom_waffle()` needs. Every square in the finished chart stands for one
cyclone.

```{r}
#| label: wstep-1
#| echo: false
#| message: false
#| warning: false
head(storms_df, 8)
```

```{r}
#| ref.label: wstep-1
#| eval: false
#| echo: true
```

### 2. Squares

`waffle::geom_waffle()` turns those counts into squares. Without faceting all
years are pooled into one block, stacked ten squares high.

```{r}
#| label: wstep-2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 150
w2 <- ggplot(storms_df, aes(fill = status, values = n)) +
  geom_waffle(color = "white", size = .25, n_rows = 10, flip = TRUE)
w2
```

```{r}
#| ref.label: wstep-2
#| eval: false
#| echo: true
```

### 3. One block per year

`facet_wrap()` with `nrow = 1` splits the squares into a column per year and
moves the year labels to the bottom, which is what makes the chart read as a
time series.

```{r}
#| label: wstep-3
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 150
w3 <- w2 +
  facet_wrap(~year, nrow = 1, strip.position = "bottom") +
  scale_x_discrete()
w3
```

```{r}
#| ref.label: wstep-3
#| eval: false
#| echo: true
```

### 4. Square squares

`coord_equal()` keeps the tiles from stretching, and the y-axis labels are
multiplied by ten so the axis counts storms rather than rows of squares.

```{r}
#| label: wstep-4
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 150
w4 <- w3 +
  scale_y_continuous(labels = function(x) x * 10, expand = c(0, 0)) +
  coord_equal()
w4
```

```{r}
#| ref.label: wstep-4
#| eval: false
#| echo: true
```

### 5. Colour

The four storm categories are coloured with the `Hiroshige` palette from
`MetBrewer`.

```{r}
#| label: wstep-5
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 5
#| fig-dpi: 150
w5 <- w4 +
  MetBrewer::scale_fill_met_d("Hiroshige", direction = 1)
w5
```

```{r}
#| ref.label: wstep-5
#| eval: false
#| echo: true
```

### 6. Text

The title states the finding outright rather than describing the chart, with the
definition of a tropical storm carried in the subtitle.

```{r}
#| label: wstep-6
#| echo: false
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 6
#| fig-dpi: 150
w6 <- w5 +
  labs(title = title_text,
       subtitle = subtitle_text,
       caption = caption_text)
w6
```

```{r}
#| ref.label: wstep-6
#| eval: false
#| echo: true
```

### 7. Theme

The last step does the most visible work: Outfit and Cabin from Google Fonts, no
axis titles, the legend moved to the top with squarish keys, and a white
background with generous margins.

```{r}
#| label: wstep-7
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 9
#| fig-dpi: 150
w6 +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(family = body_font, size = 12),
    axis.text.y = element_text(family = body_font, size = 12),
    legend.position = "top",
    legend.title = element_blank(),
    legend.spacing = unit(0.5, 'cm'),
    legend.key.height = unit(0.5, 'cm'),
    legend.key.width = unit(0.7, 'cm'),
    legend.text = element_text(family = body_font, size = 13, face = 'plain', color = "grey10"),
    plot.title.position = "plot",
    plot.title = element_textbox(margin = margin(20, 0, 10, 0), size = 30,
                                 family = title_font, face = "bold", width = unit(55, "lines")),
    plot.subtitle = element_text(margin = margin(10, 0, 20, 0), size = 16,
                                 family = body_font, color = "grey15"),
    plot.caption = element_text(family = body_font, face = "plain", size = 14,
                                color = "grey40", hjust = .5, margin = margin(20, 0, 0, 0)),
    plot.background = element_rect(color = "white", fill = "white"),
    plot.margin = margin(20, 40, 20, 40)
  )
```

```{r}
#| ref.label: wstep-7
#| eval: false
#| echo: true
```

:::
